Meta-learning in movement prediction problem of aperiodic time-series data

### 2025-01-04 doc/thesis/chapters/1.intro.tex
minor changes

\begin{itemize}
    \item Chapter \ref{chap:intro} provides the research context, problem statement, challenges and motivation.
    \item Chapter \ref{chap:related_work} presents related works, advantages and disadvantages of each study. The mentioned studies belong to two popular approaches in analyzing time-series data: Deep feature-based approach and Frequency decomposition approach. In addition, Chapter \ref{chap:related_work} also presents ML, providing the foundation for the proposed method in Chapter \ref{chap:method}.
    \item Chapter \ref{chap:method} presents a new approach for aperiodic time-series data based on the \verb|LSTM|/\verb|LSTM+CNN| feature extraction method and ML model optimization method.
    \item Chapter \ref{chap:experiment} sets up experiments, evaluation methods for the proposed method and baseline model.
    \item Chapter \ref{chap:result} analyzes the results achieved during the experiment.
    \item Chapter \ref{chap:conclusion} summarizes the main contributions of the study as well as presents future development directions.
\end{itemize}

>>>>

\begin{itemize}
    \item Chapter \ref{chap:intro} provides the research context, problem statement, challenges and motivation.
    \item Chapter \ref{chap:related_work} presents related works, advantages and disadvantages of each study. In addition, ML is also presented in this chapter, providing the foundation for the proposed method in Chapter \ref{chap:method}.
    \item Chapter \ref{chap:method} presents a new approach for aperiodic time-series data based on the \verb|LSTM|/\verb|LSTM+CNN| feature extraction method and ML model optimization method.
    \item Chapter \ref{chap:experiment} sets up experiments, evaluation methods for the proposed method and baseline model.
    \item Chapter \ref{chap:result} analyzes the results achieved during the experiment.
    \item Chapter \ref{chap:conclusion} summarizes the main contributions of the study as well as presents future development directions.
\end{itemize}

### 2025-01-04 doc/thesis/chapters/2.related_works.tex
minor changes

% % tôi sẽ chuyển cái này thành 1 idea để viết phương pháp
% % kiểu: Lấy ý tưởng từ ensemble model này kia
% \section{Ensemble models}

@@

@@

@@

@@

@@

By concatenating stacks, each receiving the remainder of the previous stack, the above architecture is expected to decompose the data into different frequency bands (weekly, daily, even hourly). In practice, \verb|NHITS| performs very well for highly periodic datasets such as electricity consumption, weather, traffic. However, we are aiming for an aperiodic time-series dataset, which has very low, or even non-existent, periodicity. This poses a huge challenge for \verb|NHITS|.

@@

One disadvantage of optimization-based ML method lies in solving equation \ref{eq:outer_opt} which requires massive overhead to compute and maintain a Hessian matrix. Even so, ML algorithms still achieve a high accuracy in handling many problem in which the quick adaptation or effective model synthesis are required.

>>>>

In this chapter, we analyze several approaches in handling time-series data, indicating the advantages and disadvantages of each approach. The related studies range from old methods such as ensemble model, deep feature-based approach, to recent methods which aim to decompose the input signal to frequency bands.

\section{Ensemble models}

% cơ chế tổng hợp cùi bắp

@@

% không có cơ chế tổng hợp multiple source

@@

% viết về conv1d, 2d, 3d...

@@

% viết 1 chút về bidirectional lstm

@@

% require massive overhead

@@

By concatenating stacks, each receiving the remainder of the previous stack, \textbf{the above architecture is expected to decompose the data into different frequency bands (weekly, daily, even hourly)}. In practice, \verb|NHITS| performs very well for highly periodic datasets such as electricity consumption, weather, traffic. \textbf{However, we are aiming for an aperiodic time-series dataset, which has very low, or even non-existent, periodicity}. This poses a huge challenge for \verb|NHITS|.

@@

\textbf{One disadvantage of optimization-based ML method lies in solving equation \ref{eq:outer_opt} which requires massive overhead to compute and maintain a Hessian matrix. Even so, ML algorithms still achieve a high accuracy in handling many problem in which the quick adaptation or effective model synthesis are required}.

### 2025-01-04 doc/thesis/chapters/3.method.tex
minor changes

We propose \verb|Temporal-ML|, a ML-based method which consists of two main parts that work in parallel: (1) - Temporal feature extraction; (2) - Models' parameter synthesis. The overview of the method is illustrated in figure \ref{fig:flow} and the detail is presented in algorithm \ref{alg:maml}. In the \textit{Temporal feature extraction} section, we use two types of features from \verb|LSTM| and \verb|LSTM+CNN| networks. In the \textit{Effective synthesis of model's parameters} section, we use \verb|MAML| to synthesize the parameters of the models.

@@

\verb|Temporal-ML| uses ML algorithms to train the model. Therefore, the data needs to be reorganized so that the ML algorithms can work. In case the data includes many different datasets belonging to the same field, each dataset will be considered a task of \verb|MAML|. In case the data includes a single dataset, it is necessary to divide this dataset into subsets corresponding to separate tasks. In summary, the prepared dataset includes $n$ tasks: $\mathcal{D} = \left\{ \mathcal{D}_t \right\}_{t=1}^{n}$. The data at each task is divided into support and query sets: $\mathcal{D}_t = \left\{ \mathcal{D}_t^{support}, \mathcal{D}_t^{query} \right\}$.

@@

The \verb|LSTM| network maintains cell-state values to selectively store long-term dependencies. This is very suitable for solving time-series data problems. On the other hand, future values often depend heavily on recent historical values. We propose to use the \verb|CNN| network to emphasize local features, thereby directing part of the model's attention to certain time points. Therefore, the proposed method not only remembers long-term features but also highlights short-term features.

@@

Hybrid ensemble models have been widely used in time-series processing problems and have been experimentally proven to be more accurate than standard time-series models because they can synthesize the strengths of many sub-models \cite{ayitey2023forex}. However, current ensemble model synthesis forms are still very rigid because they can only synthesize based on the final results (e.g. voting mechanism of bagging models) and semi-final results (e.g. stacking models). From the perspective of ensemble models, the equation \ref{eq:outer_sol} can be considered an optimization-based method of synthesizing sub-models, which helps to take advantage of the feature extraction capabilities of each model. In other words, the synthesized model can extract features at a deeper level, significantly improving the prediction ability compared to traditional ensemble models.

>>>>

We propose \verb|Temporal-ML|, a ML-based method which consists of two main components that work in parallel: (1) - Temporal feature extraction; (2) - Models' parameter synthesis. The overview of the method is illustrated in figure \ref{fig:flow} and the detail is presented in algorithm \ref{alg:maml}. In the \textit{Temporal feature extraction} section, we propose to use two types of features from \verb|LSTM| and \verb|LSTM+CNN| networks. In the \textit{Effective synthesis of model's parameters} section, \verb|MAML| is utilized to synthesize the parameters of the models.

@@

\verb|Temporal-ML| uses ML algorithms to train the model. Therefore, the data needs to be reorganized so that the ML algorithms can work. In case the data includes many different datasets belonging to the same field, each dataset will be considered a task. \textbf{The goal of using multiple datasets of the same domain is to utilize the corelation between them} as indicated in study \cite{overreactioncontrarian, mech1993portfolio}. In case the data includes a single dataset, it is necessary to divide this dataset into subsets corresponding to separate tasks. \textbf{In this way, we aim to enhance the learning of information about external factor reflected in single dataset} \cite{fama1970efficient}. In summary, the prepared dataset includes $n$ tasks: $\mathcal{D} = \left\{ \mathcal{D}_t \right\}_{t=1}^{n}$. The data at each task is divided into support and query sets: $\mathcal{D}_t = \left\{ \mathcal{D}_t^{support}, \mathcal{D}_t^{query} \right\}$.

@@

The \verb|LSTM| network maintains cell-state values to selectively store long-term dependencies. This is very suitable for solving time-series data problems. On the other hand, future values often depend heavily on recent historical values. We propose to use the \verb|CNN| network to emphasize local features, thereby directing part of the model's attention to certain time points. \textbf{Therefore, the proposed method not only remembers long-term features but also highlights short-term features. In the ideal situation, hidden constraints in aperiodic time-series data can be extracted efficiently}.

@@

Hybrid ensemble models have been widely used in time-series processing problems and have been experimentally proven to be more accurate than standard time-series models because they can synthesize the strengths of many sub-models \cite{ayitey2023forex}. However, current ensemble model synthesis forms are still very rigid because they can only synthesize based on the final results (e.g. voting mechanism of bagging models) and semi-final results (e.g. stacking models). \textbf{From the perspective of ensemble models, equation \ref{eq:outer_sol} can be considered an optimization-based method of synthesizing sub-models, which helps to take advantage of the feature extraction capabilities of each model. Moreover, equation \ref{eq:outer_sol} produces a fast-adaptive model which is extremely suitable for the context of unstable variance in data}. In other words, the synthesized model can quickly adapt to data with new variance as well as extract features at a deeper level, which significantly improves the prediction ability compared to traditional ensemble models.
